---
title: '02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Data/NYC_taxi")
```


We are now ready to use the XDF dataset for processing and analysis.  Firstly, we can start by looking at the column types, the first few rows of the data, and a summary of the `fare_amount` column.


```{r 2.01}
input_xdf <- 'yellow_tripdata_2015.xdf'
nyc_xdf <- RxXdfData(input_xdf)
rxSummary( ~ fare_amount, nyc_xdf) # provide statistical summaries for fare amount
```

Note that we could have done the same analysis with the original CSV file and skipped XDF coversion. Since we have a separate CSV file for each month, unless we combine the CSV files, we can only get the summary for one month's data.  For our purposes that will be enough.  To run `rxSummary` on the CSV file, we simply create a pointer to the CSV file using `RxTextData` (instead of `RxXdfData` as was the case above) and pass the column types directly to it.  The rest is the same.  Notice how running the summary on the CSV file takes considerably longer (even though the CSV file comprises only one month's data).


```{r 2.02}
nyc_csv <- RxTextData(input_csv, colClasses = col_classes) # point to CSV file and provide column info
rxSummary( ~ fare_amount, nyc_csv) # provide statistical summaries for fare amount
```

The last example was run to demonstrate `RevoScaleR`'s capabilty to work directly with flat files (even though they take longer than XDF files), but since our analysis involves lots of data processing and running various analytics functions, from now on we work with the XDF file, so we can benefit from faster runtime.


```{r 2.03}
rxGetInfo(nyc_xdf, getVarInfo = TRUE) # show column types and the first 10 rows
```

```{r 2.04}
head(nyc_xdf)
```
