---
title: "dplyr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("nycflights13")
library(nycflights13)
dim(flights)
head(flights)
```


##Filter rows with filter()

filter() allows you to select a subset of rows in a data frame. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame:
For example, we can select all flights on January 1st with:

```{r}
filter(flights, month == 1, day == 1)
```

This is equivalent to the more verbose code in base R:

```{r}
flights[flights$month == 1 & flights$day == 1, ]
```

filter() works similarly to subset() except that you can give it any number of filtering conditions, which are joined together with & (not && which is easy to do accidentally!). You can also use other boolean operators:

```{r}
filter(flights, month == 1 | month == 2)
```

To select rows by position, use slice():

```{r}
slice(flights, 1:10)
```


##Arrange rows with arrange()

arrange() works similarly to filter() except that instead of filtering or selecting rows, it reorders them. It takes a data frame, and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
arrange(flights, year, month, day)
```

Use desc() to order a column in descending order:

```{r}
arrange(flights, desc(arr_delay))
```


##Select columns with select()

Often you work with large datasets with many columns but only a few are actually of interest to you. select() allows you to rapidly zoom in on a useful subset using operations that usually only work on numeric variable positions:

```{r}
# Select columns by name
select(flights, year, month, day)
```

```{r}
# Select all columns between year and day (inclusive)
select(flights, year:day)
```

```{r}
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```


##Extract distinct (unique) rows

Use distinct()to find unique values in a table:

```{r}
distinct(flights, tailnum)
```


##Add new columns with mutate()

Besides selecting sets of existing columns, it’s often useful to add new columns that are functions of existing columns. This is the job of mutate():

```{r}
mutate(flights, gain = arr_delay - dep_delay, speed = distance / air_time * 60)
```

If you only want to keep the new variables, use transmute():
```{r}
transmute(flights, gain = arr_delay - dep_delay, gain_per_hour = gain / (air_time / 60)
)
```


##Summarize values with summarise()

The last verb is summarise(). It collapses a data frame to a single row:

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```


##Grouped operations

These verbs are useful on their own, but they become really powerful when you apply them to groups of observations within a dataset. In dplyr, you do this by with the group_by() function. It breaks down a dataset into specified groups of rows. When you then apply the verbs above on the resulting object they’ll be automatically applied “by group”. Most importantly, all this is achieved by using the same exact syntax you’d use with an ungrouped object.

In the following example, we split the complete dataset into individual planes and then summarize each plane by counting the number of flights (count = n()) and computing the average distance (dist = mean(Distance, na.rm = TRUE)) and arrival delay (delay = mean(ArrDelay, na.rm = TRUE)). We then use ggplot2 to display the output.

```{r}
library(ggplot2)
by_tailnum <- group_by(flights, tailnum)
delay <- summarise(by_tailnum,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE))
delay <- filter(delay, count > 20, dist < 2000)

# Interestingly, the average delay is only slightly related to the
# average distance flown by a plane.
ggplot(delay, aes(dist, delay)) +
  geom_point(aes(size = count), alpha = 1/2) +
  geom_smooth() +
  scale_size_area()
```
